---
title: "A (very brief) introduction to equivalence testing"
author: Owen
date: "2026-04-09"
date-format: long
df-print: kable
format: 
  revealjs:
    logo: logo.svg
    footer: "Introduction to equivalence testing"
---

##  {#slide1-id data-menu-title="Slide 1"}

```{css, echo = FALSE}
.justify {
  text-align: center
}
```

```{r, echo = FALSE}
library(ggplot2)
library(dplyr)
library(magrittr)
library(tibble)
library(TOSTER)
library(brms)
library(bayestestR)

generate_example_data <- function(n_per_group, mean_wt, mean_mutant, sd) {
  data <- tibble(
    Genotype = rep(c("Wild type", "Mutant"), each = n_per_group),
    Value = c(
      rnorm(n = n_per_group, mean = mean_wt, sd = sd),
      rnorm(n = n_per_group, mean = mean_mutant, sd = sd)
    )
  )
  
  data$Genotype %<>% factor() %>% relevel("Wild type")
  
  data
}

add_p_val <- function(plot, y_pos, lab, col="black") {
  plot +
    annotate("segment", x = 1, xend = 2, y = y_pos, yend = y_pos, color = col) +
    annotate("segment", x = 1, xend = 1, y = y_pos - 0.2, yend = y_pos, color = col) +
    annotate("segment", x = 2, xend = 2, y = y_pos - 0.2, yend = y_pos, color = col) +
    annotate("text", x = 1.5, y = y_pos + 0.4, label = lab, size = 7, color = col)
}
```

:::{.justify}
```{r, fig.width=7, fig.height=7} 
set.seed(42)

MEAN_WT <- 10
MEAN_MUTANT <- 11.4

genotype_example_data <- generate_example_data(24, MEAN_WT, MEAN_MUTANT, 1.5)

genotype_t_test <- t.test(Value ~ Genotype, data = genotype_example_data)
p_value_label <- paste0("p = ", format.pval(genotype_t_test$p.value, digits = 2, eps = 0.001))
y_position <- max(genotype_example_data$Value) + 0.6

basic_plot <- genotype_example_data %>%
  ggplot(aes(x = Genotype, y = Value)) +
  geom_point(
    position = position_jitter(width = 0.12, height = 0, seed = 42),
    size = 2,
    alpha = 0.85
  ) +
  xlab(NULL) +
  ylab("Very important variable") +
  theme_classic() +
  theme(
    axis.text = element_text(size = 18),
    axis.title = element_text(size = 24)
  )

basic_plot_p_vals <- basic_plot %>% add_p_val(y_position, p_value_label)
basic_plot_p_vals %>% print()
```
:::

## {#slide2-id data-menu-title="Slide 2"}

::: {.incremental}
- **Scientist**: There is no difference between the wild type and mutant genotypes.
- **Reviewer**: But you haven't really shown that. What you have shown is that _assuming the null hypothesis of no difference between genotypes is true_, you would only expect to see a difference as extreme as the one you observed (or more extreme) 7.5% of the time. That's not the same as showing there is no difference between genotypes.
- **Scientist**: ☹️
:::

##  {#slide3-id data-menu-title="Slide 3"}

:::{.justify}
```{r, fig.width=7, fig.height=7} 
basic_plot_p_vals %>% print()
```
:::

##  {#slide3a-id data-menu-title="Slide 3a"}

:::{.justify}
```{r, fig.width=7, fig.height=7}
additional_genotype_example_data <- generate_example_data(96, MEAN_WT, MEAN_MUTANT, 1.5) %>% 
  rbind(genotype_example_data) 

genotype_t_test <- t.test(Value ~ Genotype, data = additional_genotype_example_data)
p_value_label <- paste0("p = ", format.pval(genotype_t_test$p.value, digits = 2, eps = 0.001))
y_position <- max(additional_genotype_example_data$Value) + 0.6

additional_data_plot <- additional_genotype_example_data %>%
  ggplot(aes(x = Genotype, y = Value)) +
  geom_point(
    position = position_jitter(width = 0.12, height = 0, seed = 42),
    size = 2,
    alpha = 0.85
  ) +
  xlab(NULL) +
  ylab("Very important variable") +
  theme_classic() +
  theme(
    axis.text = element_text(size = 18),
    axis.title = element_text(size = 24)
  )

additional_data_plot %>% add_p_val(y_position, p_value_label, col = "white") %>% print()
```
:::

##  {#slide3b-id data-menu-title="Slide 3b"}

:::{.justify}
```{r, fig.width=7, fig.height=7}
additional_data_plot %<>% add_p_val(y_position, p_value_label) 

additional_data_plot %>% print()
```
:::

##  {#slide4-id data-menu-title="Slide 4"}

:::{.justify}
```{r, fig.width=7, fig.height=7}
additional_data_plot <- additional_data_plot + 
  annotate("segment", x = 0.8, xend = 1.2, y = MEAN_WT, yend = MEAN_WT, colour = "red", linewidth = 1.2) +
  annotate("segment", x = 1.8, xend = 2.2, y = MEAN_MUTANT, yend = MEAN_MUTANT, colour = "red", linewidth = 1.2)
 
additional_data_plot %>% print()
```
:::

##  {#slide5-id data-menu-title="Slide 5"}

:::{.justify}
```{r, fig.width=7, fig.height=7}
additional_data_plot <- additional_data_plot + ylim(0, 16)

additional_data_plot %>% print()
```
:::

##  {#slide6-id data-menu-title="Slide 6"}

::: {.r-fit-text .absolute top=200}
Is this difference small enough to be of no _practical_ significance?
:::

##  {#slide7-id data-menu-title="Slide 7"}

::: {.incremental}
- Q: Could the true difference be bigger than what we’d consider negligible?
  - "TOST": "two one-sided t-tests"
  - frequentist
- Q: Given the data, how likely is it that the difference is negligible?
  - "ROPE": "region of practical equivalence"
  - Bayesian (😱)
- in _either_ case, we must define what counts as "no practical difference"
:::

##  {#slide8-id data-menu-title="Slide 8"}

:::{.justify}
```{r, fig.width=7, fig.height=7}
basic_plot %>% print()
```
:::

##  {#slide9-id data-menu-title="Slide 9"}

:::{.justify}
```{r, fig.width=7, fig.height=7}
real_mean_wt <- genotype_example_data %>% filter(Genotype == "Wild type") %>% pull(Value) %>% mean()

basic_plot_equivalence <- basic_plot +
  annotate("segment", x = 0.8, xend = 1.2, y = real_mean_wt, yend = real_mean_wt, colour = "red", linewidth = 1.2)

basic_plot_equivalence %>% print()
```
:::

##  {#slide10-id data-menu-title="Slide 10"}

:::{.justify}
```{r, fig.width=7, fig.height=7}
basic_plot_equivalence <- basic_plot_equivalence +
  geom_hline(yintercept = real_mean_wt + 2, colour = "red", linetype = "dashed") +
  geom_hline(yintercept = real_mean_wt - 2, colour = "red", linetype = "dashed")

basic_plot_equivalence %>% print()
```
:::

##  {#slide11-id data-menu-title="Slide 11"}

:::{.justify}
```{r, fig.width=7, fig.height=7}
real_mean_mutant <- genotype_example_data %>% filter(Genotype == "Mutant") %>% pull(Value) %>% mean()

basic_plot_equivalence <- basic_plot_equivalence +
  annotate("segment", x = 1.8, xend = 2.2, y = real_mean_mutant, yend = real_mean_mutant, colour = "blue", linewidth = 1.2)

basic_plot_equivalence %>% print()
```
:::

##  {#slide12-id data-menu-title="Slide 12"}

```{r}
t_TOST(formula = Value ~ Genotype,
       data = genotype_example_data,
       paired = TRUE,
       eqb = 2) %>% print()
```

##  {#slide13-id data-menu-title="Slide 13"}

::: {.r-fit-text .absolute top=200}
"Using the paired t-test, a null hypothesis significance test (NHST), and an equivalence test, via two one-sided tests (TOST), were performed with an alpha-level of 0.05. These tested the null hypotheses that true mean difference is equal to 0 (NHST), and true mean difference is more extreme than -2 and 2 (TOST). The equivalence test was significant, t(23) = 2.236, p = 0.018 (mean difference = -0.925 90% C.I.[-1.75, -0.102]). At the desired error rate, it can be stated that the true mean difference is between -2 and 2."
:::

##  {#slide14-id data-menu-title="Slide 14"}

::: {.incremental}
- Under the null hypothesis that the true difference is greater than 2, you would expect to see a difference (Mutant vs Wild type) as small in magnitude as the one that you have observed (or smaller) around 2% of the time.
- Under the null hypothesis that the true difference is less than -2, you would expect to see a difference (Mutant vs Wild type) as large in magnitude as the one that you have observed (or larger) less than 0.1% of the time.
- Under the null hypothesis of _no_ difference you would expect to see a difference as extreme as the one you observed (or more extreme) 7.5% of the time.
:::

##  {#slide15-id data-menu-title="Slide 15"}

::: {.r-fit-text .absolute top=200}
How likely is it that the difference is negligible?
:::

##  {#slide16-id data-menu-title="Slide 16"}

:::{.justify}
```{r, fig.width=9, fig.height=7}
prior_distribution_data <- tibble(
  mean_difference = seq(-20, 20, length.out = 400),
  density = dnorm(mean_difference, mean = 0, sd = 5)
)

prior_distribution_data %>%
  ggplot(aes(x = mean_difference, y = density)) +
  geom_line(linewidth = 1.2) +
  ylim(0, 0.4)+
  xlab("Mutant vs Wild type mean difference") +
  ylab("Density") +
  theme_classic() +
  theme(
    axis.text = element_text(size = 18),
    axis.title = element_text(size = 24)
  )
```
:::

##  {#slide17-id data-menu-title="Slide 17"}

:::{.justify}
```{r, fig.width=9, fig.height=7}
posterior_distribution_data <- tibble(
  mean_difference = seq(-20, 20, length.out = 400),
  density = dnorm(mean_difference, mean = 1.4, sd = 1)
)

prior_posterior_plot <- prior_distribution_data %>% ggplot(aes(x = mean_difference, y = density)) +
  geom_line(linewidth = 1.2, colour="grey") +
  geom_line(data=posterior_distribution_data, linewidth = 1.2) +
  ylim(0, 0.4)+
  xlab("Mutant vs Wild type mean difference") +
  ylab("Density") +
  theme_classic() +
  theme(
    axis.text = element_text(size = 18),
    axis.title = element_text(size = 24)
  )

prior_posterior_plot %>% print()
```
:::

##  {#slide18-id data-menu-title="Slide 18"}

:::{.justify}
```{r, fig.width=9, fig.height=7}
prior_posterior_plot <- prior_posterior_plot +
  geom_vline(xintercept = -2, colour = "red", linetype = "dashed") +
  geom_vline(xintercept = 2, colour = "red", linetype = "dashed")

prior_posterior_plot %>% print()
```
:::

##  {#slide18a-id data-menu-title="Slide 18a"}
:::{.justify}
```{r, fig.width=9, fig.height=7}
prior_posterior_plot <- prior_posterior_plot +
  geom_area(
    data = posterior_distribution_data %>% filter(mean_difference >= -2, mean_difference <= 2),
    aes(x = mean_difference, y = density),
    inherit.aes = FALSE,
    fill = "grey",
    alpha = 0.4
  )

prior_posterior_plot %>% print()
```
:::

##  {#slide19-id data-menu-title="Slide 19"}

```{r include=FALSE}
bm <- brm(
  Value ~ Genotype,
  data = genotype_example_data,
  prior = prior(normal(0, 5), class = "b"),
  iter = 10000,
)
```

```{r}
posterior_summary(bm)
```

##  {#slide20-id data-menu-title="Slide 20"}

:::{.justify}
```{r, fig.width=9, fig.height=7}
post <- posterior_samples(bm, pars = "b_GenotypeMutant")
density_estimate <- density(post$b_GenotypeMutant, n = 1024)

density_df <- tibble(
  b_GenotypeMutant = density_estimate$x,
  density = density_estimate$y
)

rope_low  <- -2
rope_high <-  2

posterior_draws_plot <- ggplot(density_df, aes(x = b_GenotypeMutant, y = density)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "solid", colour = "grey") +
  labs(
    x = "Mutant vs Wild type mean difference",
    y = "Density"
  ) +
  xlim(-2.5, 2.5) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 18),
    axis.title = element_text(size = 24)
  )

posterior_draws_plot %>% print()
```
:::

##  {#slide21-id data-menu-title="Slide 21"}

:::{.justify}
```{r, fig.width=9, fig.height=7}
posterior_draws_plot <- posterior_draws_plot +
  geom_area(
    data = density_df %>% filter(b_GenotypeMutant >= rope_low, b_GenotypeMutant <= rope_high),
    aes(x = b_GenotypeMutant, y = density),
    inherit.aes = FALSE,
    fill = "grey",
    alpha = 0.3
  ) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = c(rope_low, rope_high), linetype = "dashed", color = "red")

posterior_draws_plot %>% print()
```
:::

##  {#slide22-id data-menu-title="Slide 22"}

```{r}
p_rope(post$b_GenotypeMutant, range = c(-2, 2)) %>% print()
```

##  {#slide22a-id data-menu-title="Slide 22a"}

::: {.r-fit-text .absolute top=200}
"The posterior mean (of the genotype effect) is approximately 0.92 with a 95% credible interval of approximately -0.08 to 1.92. ROPE
analysis indicates that approximately 98.3% of the posterior distribution for the genotype effect falls within the equivalence bounds of
-2 to 2."
:::

##  {#slide23-id data-menu-title="Slide 23"}

:::{.justify}
```{r, fig.width=7, fig.height=7}
basic_plot %>% print()
```
:::
